White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-12-13 papers/wdmerger_I/paper.tex
Update of the paper with latest numerical results

We now consider the phase of the binary system where the stars are orbiting each other 
at distances great enough that the initial orbits should be approximately Keplerian. 
There are a number of effects worth looking into here. For simplicity, we choose two 
cases to demonstrate the simulation behavior: an equal mass case of two $0.9\ \msolar$ 
white dwarfs, and an unequal mass case of $0.9\ \msolar$ and $0.6\ \msolar$ white dwarfs.
In each case, the initial orbital period is 100 seconds.

@@

We are most interested in the stability of these systems over long
timescales. To this end, we consider the same systems as above, but
evolve them for 25 orbital periods. In
\autoref{fig:circular_orbit_comparison} we illustrate the evolution of
these systems by plotting the center of mass locations of the white
dwarfs on the orbital ($xy$) plane. For the equal mass case in the
inertial reference frame, the curves essentially fall on top of each
other, indicating that the stars are indeed orbiting at the initial
distance without substantial deviation. In the unequal mass case, the
stars now orbit at different distances but do not much deviate from
their initial distance. We do observe a secular decline in the center
of mass of these stars, indicating that eventually the stellar orbits
will degrade due to numerical error even without intervention. The
co-rotating frame is different.  For clarity of visualization, we
rotate these results back into the inertial frame before displaying
their orbits.  The stellar orbit in the equal mass case does not show
the same secular decline, and the system appears to be
stable. The cost is that the oscillations about this
equilibrium are slightly larger than in the inertial frame. The
unequal mass case displays this effect starkly, to the point where the system 
becomes unstable: while the average distance is not substantially declining, 
there are huge oscillations about this
average. In addition, the stars move substantially within the rotating
frame itself, yielding the chaotic-looking orbits in the figure. The
oscillations grow with time and will eventually cause the stars to
merge; however, the stars would eventually merge even in the inertial
frame case, and at a faster rate than the equal mass case. This
suggests that this mass ratio is too large to ignore tidal effects,
and that the tidal effects play out much differently in the rotating
frame.

@@

Turning to the conservation properties of the system, we examine as
fairly typical cases the equal mass system in the inertial frame for
energy conservation (\autoref{fig:energy_conservation_equal}), and the
unequal mass system in the rotating frame for angular momentum
conservation (\autoref{fig:angular_momentum_conservation_unequal}).
Angular momentum is conserved to within a few percent over the 25
orbits, while energy conservation is about an order of magnitude
better. We note that while this is already a fairly good level of
energy conservation, it is not nearly as good as the results of
\citet{marcello:2012}. This is because we reset the internal energy to
a level corresponding to our temperature floor when it goes negative,
while \citeauthor{marcello:2012} do not reset and instead ignore the
internal energy if it is negative. The resets impose an artificial
floor on our ability to conserve energy, but they only happen in
low-density regions and do not much affect the large-scale dynamics.
Meanwhile, relative angular momentum conservation is not quite as good 
as relative energy conservation.  This is
linked to the decline (or increase) in the size of the orbit. This
implies that we ought to be careful in concluding that at these
moderate resolutions we can safely evolve systems for many dozens of
orbits; this needs to be verified to ensure that an observed inspiral
and merger is physically (not numerically) motivated.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/unequal_angular_momentum_rot1.eps}
  \caption{Absolute magnitude of the relative change in angular momentum of two equal mass white dwarfs after 25 orbital periods, 
           evolved in a co-rotating reference frame. We consider only the component of the angular moment along the rotational 
           axis. The decline and recovery is a change in sign of the angular momentum difference.
           \label{fig:angular_momentum_conservation_unequal}}
\end{figure}

As a simple verification test to ensure our gravitational wave calculations are correct, we plot the 
gravitational wave strain along the rotation axis for the first two periods of the unequal mass system. 
At this early time the orbit is circular and so to a good approximation we expect that the gravitational 
wave signal should be that of two point masses, whose positions are:
\begin{align}
  \mathbf{r}_P(t) &= -a_P\, \text{cos}(\omega t) \hat{x} - a_P\, \text{sin}(\omega t) \hat{y} \\
  \mathbf{r}_S(t) &= a_S\, \text{cos}(\omega t) \hat{x} + a_S\, \text{sin}(\omega t) \hat{y}.
\end{align}
Then the mass distribution is $\rho(\mathbf{r}) = M_P\, \delta^3(\mathbf{r} - \mathbf{r}_P) + M_S\, \delta^3(\mathbf{r} - \mathbf{r}_S)$.
From this it is straightforward to calculate the quadruopole tensor, take its second time derivative, and then apply the 
projection operator to get the gravitational wave polarizations along the rotation axis:
\begin{align}
  h_+ &= -4\frac{G\mu}{c^4 r}\left[G M_{\text{tot}} \omega \right]^{2/3}\, \text{cos}(2\omega t) \\
  h_\times &= -4\frac{G\mu}{c^4 r}\left[G M_{\text{tot}} \omega \right]^{2/3}\, \text{sin}(2\omega t).
\end{align}
$\mu$ is the reduced mass, while $M_{\text{tot}}$ is the total mass. From this we see that the 
gravitational wave frequency is twice the orbital frequency, and that the two polarizations 
are out of phase by $90^\circ$ in time. We compare this analytical expectation to the 
numerical results in \autoref{fig:gw_strain}. We find very good agreement in this case, and this 
level of agreement holds in the rotating frame as well.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/unequal_gw_rot0.eps}
  \caption{Gravitational wave strain polarizations for the first two orbital periods of the 
           unequal mass system. The curves with markers are the numerical data, while the 
           curves without markers are the analytical results for two point masses.\label{fig:gw_strain}}
\end{figure}

Finally we consider whether the dynamical behavior of the system converges with resolution. 
In \autoref{fig:unequal_spatial_convergence} we plot the first 10 seconds of the orbit for 
the unequal mass system, at three different resolutions in the inertial frame: our default 
resolution of $256^3$ zones, as well as a single level of refinement with a jump by a factor of 
two (effective resolution $512^3$) or a jump by a factor of four (effective resolution $1024^3$). 
It is clear that at the latter resolution (corresponding to physical resolution of 100 km), 
we have achieved convergent behavior. In the rotating frame (not shown here), the results are 
not quite the same but for these resolutions the results do tend toward convergence.

@@

We presented a set of numerical tests that show where we can 
and cannot trust these techniques. We spent much time considering 
the role of bulk motions on the grid and we conclude that there 
are real issues with substantial bulk velocities on static grids, 
that can diminish the quality of the resulting solutions to the 
fluid equations, but that these effects diminish with increasing 
resolution. We therefore make no explicit claims about the usefulness 
of Langrangian versus Eulerian methods and instead simply observe 
that whenever a simulation is performed, it is important to have 
a measure of numerical accuracy and a sense of whether we are 
witnessing properties that converge with increasing resolution.
These problems do suggest that, where possible, we should seek to 
minimize bulk motions on static grids. A comparison of orbit 
simulations in both rotating and inertial reference frames 
demonstrates that in practice this is not so simple, and that 
a rotating reference frame has its own numerical issues for 
the type of simulation we desire to perform here. Based on 
the tests performed here, in some cases the stars may be 
\textit{more} stable in an inertial frame despite 
constantly being in motion on the grid. It 
is not easy to predict the correct behavior of such systems, 
though one recourse for assessing confidence in a model from a 
verification standpoint is to see whether the observed 
behavior converges with resolution. It is not clear 
whether at practical resolutions the results in the 
rotating and inertial frames will converge to each other.

>>>>

We now consider the phase of the binary system where the stars are orbiting each other 
at distances great enough that the initial orbits should be approximately Keplerian. 
There are a number of effects worth looking into here. For simplicity, we choose two 
cases to demonstrate the simulation behavior: an equal mass case of two $0.9\ \msolar$ 
white dwarfs, and an unequal mass case of $0.9\ \msolar$ and $0.75\ \msolar$ white dwarfs.
In both systems, the secondary should be stable against mass loss.
In each case, the initial orbital period is 100 seconds.

@@

We are most interested in the stability of these systems over long
timescales. To this end, we consider the same systems as above, but
evolve them for 25 orbital periods. In
\autoref{fig:circular_orbit_comparison} we illustrate the evolution of
these systems by plotting the center of mass locations of the white
dwarfs on the orbital ($xy$) plane. For the equal mass case in the
inertial reference frame, the curves fall nearly on top of each
other for most of the run, indicating that the stars are indeed 
orbiting at the initial distance, at least for a while. Towards the end 
of the run, however, the orbit starts to decay significantly, and the center-of-mass
distance of the two stars has decreased by about 10\% after 25 orbits.
We attribute this to non-conservation of angular momentum, which occurs 
because our code only explicitly conserves linear momentum. This orbital 
decay resembles the effect seen by \citet{swc:2000} for the case of neutron stars. 
In the unequal mass case, the magnitude of the orbital decay is smaller but 
at the end of the run the secular decline in distance is also visible. In 
both cases the stars would likely merge due to numerical error 
after a long enough timescale.

The co-rotating frame is different.  For clarity of visualization, we
rotate these results back into the inertial frame before displaying
their orbits.  In both the equal and unequal mass cases, the centrifugal force 
pushes the stars outward toward a new equilibrium distance that is a few 
percent larger than its initial distance. At the end of the run, the system is 
relatively stable, with oscillations about the new equilibrium distance. In fact 
these oscillations occur too in the inertial frame, but they are much more pronounced 
here. In the unequal mass case this is coupled with severe precession of the orbit, 
which results in chatoic-looking orbits when viewed from the rotating reference frame. 
These result from the explicit numerical consideration of the Coriolis and centrifugal 
terms, which do not appear in the inertial frame. So while the rotating frame 
is clearly more stable than the inertial frame, the cost is that the specific 
dynamics may be more suspect.

@@

Turning to the conservation properties of the system, we examine as
fairly typical cases the equal mass system in the inertial frame for
energy conservation (\autoref{fig:energy_conservation_equal}), and the
unequal mass system in the rotating frame for angular momentum
conservation (\autoref{fig:angular_momentum_conservation_unequal}).
Angular momentum is conserved to within 10 percent over the 25
orbits, while energy conservation is about an order of magnitude
better. We note that while this is already a fairly good level of
energy conservation, it is not nearly as good as the results of
\citet{marcello:2012}. This is because we reset the internal energy to
a level corresponding to our temperature floor when it goes negative,
while \citeauthor{marcello:2012} do not reset and instead ignore the
internal energy if it is negative. The resets impose an artificial
floor on our ability to conserve energy, but they only happen in
low-density regions and do not much affect the large-scale dynamics.
Meanwhile, relative angular momentum conservation is not quite as good 
as relative energy conservation.  This is 
linked to the decline (or increase) in the size of the orbit. This
implies that we ought to be careful in concluding that at these
moderate resolutions we can safely evolve systems for many dozens of
orbits; this needs to be verified to ensure that an observed inspiral
and merger is physically (not numerically) motivated.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/unequal_angular_momentum_rot1.eps}
  \caption{Absolute magnitude of the relative change in angular momentum of two unequal mass white dwarfs after 25 orbital periods, 
           evolved in a co-rotating reference frame. We consider only the component of the angular moment along the rotational 
           axis.
           \label{fig:angular_momentum_conservation_unequal}}
\end{figure}

As a simple verification test to ensure our gravitational wave calculations are correct, we plot the 
gravitational wave strain along the rotation axis for the first two periods of an unequal mass system. 
At this early time the orbit is circular and so to a good approximation we expect that the gravitational 
wave signal should be that of two point masses, whose positions are:
\begin{align}
  \mathbf{r}_P(t) &= -a_P\, \text{cos}(\omega t) \hat{x} - a_P\, \text{sin}(\omega t) \hat{y} \\
  \mathbf{r}_S(t) &= a_S\, \text{cos}(\omega t) \hat{x} + a_S\, \text{sin}(\omega t) \hat{y}.
\end{align}
Then the mass distribution is $\rho(\mathbf{r}) = M_P\, \delta^3(\mathbf{r} - \mathbf{r}_P) + M_S\, \delta^3(\mathbf{r} - \mathbf{r}_S)$.
From this it is straightforward to calculate the quadruopole tensor, take its second time derivative, and then apply the 
projection operator to get the gravitational wave polarizations along the rotation axis:
\begin{align}
  h_+ &= -4\frac{G\mu}{c^4 r}\left[G M_{\text{tot}} \omega \right]^{2/3}\, \text{cos}(2\omega t) \\
  h_\times &= -4\frac{G\mu}{c^4 r}\left[G M_{\text{tot}} \omega \right]^{2/3}\, \text{sin}(2\omega t).
\end{align}
$\mu$ is the reduced mass, while $M_{\text{tot}}$ is the total mass. From this we see that the 
gravitational wave frequency is twice the orbital frequency, and that the two polarizations 
are out of phase by $90^\circ$ in time. We compare this analytical expectation to the 
numerical results in \autoref{fig:gw_strain}. We find very good agreement in this case, and this 
level of agreement holds in the rotating frame as well.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/unequal_gw_rot0.eps}
  \caption{Gravitational wave strain polarizations for the first two orbital periods of an 
           unequal mass system. The curves with markers are the numerical data, while the 
           curves without markers are the analytical results for two point masses.\label{fig:gw_strain}}
\end{figure}

Finally we consider whether the dynamical behavior of the system converges with resolution. 
In \autoref{fig:unequal_spatial_convergence} we plot the first 10 seconds of the orbit for 
the unequal mass system, at three different resolutions in the inertial frame: our default 
resolution of $256^3$ zones, as well as a single level of refinement with a jump by a factor of 
two (effective resolution $512^3$) or a jump by a factor of four (effective resolution $1024^3$). 
It is clear that at the latter resolution (corresponding to physical resolution of 100 km), 
we have achieved convergent behavior. In the rotating frame (not shown here), the results are 
not quite the same but for these resolutions the results do tend toward convergence. These 
results suggest a minimum resolution of 200 km for getting the dynamics right.

@@

We presented a set of numerical tests that show where we can 
and cannot trust these techniques. We spent much time considering 
the role of bulk motions on the grid and we conclude that there 
are real issues with substantial bulk velocities on static grids, 
that can diminish the quality of the resulting solutions to the 
fluid equations, but that these effects diminish with increasing 
resolution. We therefore make no explicit claims about the usefulness 
of Langrangian versus Eulerian methods and instead simply observe 
that whenever a simulation is performed, it is important to have 
a measure of numerical accuracy and a sense of whether we are 
witnessing properties that converge with increasing resolution.
These problems do suggest that, where possible, we should seek to 
minimize bulk motions on static grids. A comparison of orbit 
simulations in both rotating and inertial reference frames 
demonstrates that in practice this is not so simple, and that 
a rotating reference frame has its own numerical issues for 
the type of simulation we desire to perform here. While decay 
of the stellar orbit is a commonplace feature in the inertial frame 
but easier to avoid in the rotating frame, the rotation forces 
can result in likely unphysical oscillations of the stars. It 
is not easy to predict the correct behavior of such systems,
and therefore determine which frame is closer to being correct, 
though one recourse for assessing confidence in a model from a 
verification standpoint is to see whether the observed 
behavior converges with resolution. It is not clear 
whether at practical resolutions the results in the 
rotating and inertial frames will converge to each other.
